---
title: "Stats Inflation Analysis"
author: "Derek Fu"
date: "`r Sys.Date()`"
output: html_document
---


```{r message=FALSE}

source("main/data_extraction.R")
source("main/gibbs_sampler.R")

library(coda)
library(ggplot2)

set.seed(123)

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "D:/NBA-Stats-Inflation/src")
```


## Poisson-Gamma Model on Total Points of Top Scorers:  

1. First, load the data and call the necessary functions to clean it:

```{r message=FALSE}

player_data = read_csv("data/player_stats.csv")

top_10_scorers = get_top_n_scorers(player_data, 10)
top_10_sum = get_top_scorers_sum(top_10_scorers)

head(top_10_sum)

```

2. Modify some rows due to different numbers of games played in certain seasons:

```{r}

top_10_sum$Total_PTS[top_10_sum$Season == 1999] = as.integer(top_10_sum$Total_PTS[top_10_sum$Season == 1999] * (82/50))
top_10_sum$Total_PTS[top_10_sum$Season == 2012] = as.integer(top_10_sum$Total_PTS[top_10_sum$Season == 1999] * (82/66))
top_10_sum$Total_PTS[top_10_sum$Season == 2020] = as.integer(top_10_sum$Total_PTS[top_10_sum$Season == 1999] * (82/71))
top_10_sum$Total_PTS[top_10_sum$Season == 2021] = as.integer(top_10_sum$Total_PTS[top_10_sum$Season == 1999] * (82/72))
top_10_sum$Total_PTS[top_10_sum$Season == 2024] = as.integer(top_10_sum$Total_PTS[top_10_sum$Season == 1999] * (82/55))

top_10_sum$Total_PTS = as.integer(top_10_sum$Total_PTS / 100)

```

Also, check the mean and variance for Poisson assumption:

```{r}

mean(top_10_sum$Total_PTS); var(top_10_sum$Total_PTS)

```


3. Plot the data:

```{r}

# top_10_sum$Season = factor(top_10_sum$Season)

ggplot(top_10_sum, aes(x = as.factor(Season), y = Total_PTS)) +
  geom_point(color = "blue", size = 2) + 
  labs(
    title = "Total Points Scored by Top 10 NBA Scorers Each Season",
    x = "Season",
    y = "Total Points (in hundreds)",
    caption = "Data Source: BBR"
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

4. Run the models:

```{r}

# result = run_poisson_model(top_10_sum$Total_PTS, alpha = 0.5, beta = 0.01)
# posterior_changepoint = result[[3]]
# hist(posterior_changepoint + 1977, breaks = 50)

result = run_poisson_model(top_10_sum$Total_PTS[-48], alpha = 0.5, beta = 0.01)
posterior_changepoint = result[[3]]
ci = HPDinterval(as.mcmc(posterior_changepoint + 1977))

# Plot histogram of potential change-points:
hist(posterior_changepoint + 1977, breaks = 50, col = "skyblue", border = "black",
     main = "Histogram of Potential Changepoints",
     xlab = "Potential Change-point Season in Top Scorers",
     ylab = "Frequency")

# Add the credible intervals and the corresponding legend:
abline(v = ci, lwd = 3, col = 'blue')
legend("topright", legend = "95% Credible Interval", col = "blue", lwd = 3, cex = 0.5)

```

```{r}

ci = HPDinterval(as.mcmc(posterior_changepoint + 1977))
mean_posterior = round(mean(posterior_changepoint) + 1977)

plot(top_10_sum$Season[-48], top_10_sum$Total_PTS[-48],
     main = 'Total Points by Top 10 NBA Scorers (Change-point in Late 1990s)',
     xlab = 'Season', ylab = 'Total Points', col = 'black', pch = 16)

points(top_10_sum$Season[-48], top_10_sum$Total_PTS[-48], col = 'black', pch = 19)
abline(v = ci, col = 'blue', lwd = 2)
abline(v = mean_posterior, col = 'red', lwd = 2)

# Add legend
legend("topright", legend = c("Total Points", "95% Credible Interval", "Posterior Mean"),
       col = c("black", "blue", "red"), pch = c(19, NA, NA), lwd = 2, cex = 0.5,
       title = "Legend")

```


5. Check convergence: 

```{r}

plot(as.ts(posterior_changepoint))

```

```{r}

effectiveSize(posterior_changepoint)

```

## Poisson-Gamma Model on Total Number of Team Fouls:

```{r}

team_per_game_data = read_csv("data/team_per_game_stats.csv")

top_10_team_fouls = get_top_n_team_fouls(team_per_game_data, 10)
top_10_foul_sum = get_team_foul_sum(top_10_team_fouls)

head(top_10_foul_sum)

```

Similarly, check mean and variance after some modifications:  

```{r}

top_10_foul_sum$Total_Fouls[top_10_foul_sum$Season == 1999] = as.integer(top_10_foul_sum$Total_Fouls[top_10_foul_sum$Season == 1999] * (82/50))
top_10_foul_sum$Total_Fouls[top_10_foul_sum$Season == 2012] = as.integer(top_10_foul_sum$Total_Fouls[top_10_foul_sum$Season == 1999] * (82/66))
top_10_foul_sum$Total_Fouls[top_10_foul_sum$Season == 2020] = as.integer(top_10_foul_sum$Total_Fouls[top_10_foul_sum$Season == 1999] * (82/71))
top_10_foul_sum$Total_Fouls[top_10_foul_sum$Season == 2021] = as.integer(top_10_foul_sum$Total_Fouls[top_10_foul_sum$Season == 1999] * (82/72))
top_10_foul_sum$Total_Fouls[top_10_foul_sum$Season == 2024] = as.integer(top_10_foul_sum$Total_Fouls[top_10_foul_sum$Season == 1999] * (82/55))

top_10_foul_sum$Total_Fouls = as.integer(top_10_foul_sum$Total_Fouls / 100)

```

```{r}

mean(top_10_foul_sum$Total_Fouls); var(top_10_foul_sum$Total_Fouls)

```

```{r}

ggplot(top_10_foul_sum, aes(x = as.factor(Season), y = Total_Fouls)) +
  geom_point(color = "blue", size = 2) + 
  labs(
    title = "Total Team Fouls by Top 10 Teams in Fouls Each Season",
    x = "Season",
    y = "Total Points (in hundreds)",
    caption = "Data Source: BBR"
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}

result = run_poisson_model(top_10_foul_sum$Total_Fouls[-48], alpha = 0.5, beta = 0.01)
posterior_changepoint = result[[3]]
ci = HPDinterval(as.mcmc(posterior_changepoint + 1977))

# Plot histogram of potential change-points:
hist(posterior_changepoint + 1977, breaks = 50, col = "skyblue", border = "black",
     main = "Histogram of Potential Changepoints",
     xlab = "Potential Change-point Season",
     ylab = "Frequency")

# Add the credible intervals and the corresponding legend:
abline(v = ci, lwd = 3, col = 'blue')
legend("topright", legend = "95% Credible Interval", col = "blue", lwd = 3, cex = 0.5)

```

```{r}

ci = HPDinterval(as.mcmc(posterior_changepoint + 1977))
mean_posterior = round(mean(posterior_changepoint) + 1977)

plot(top_10_foul_sum$Season[-48], top_10_foul_sum$Total_Fouls[-48],
     main = 'Total Fouls by Top 10 Teams in Fouls (Change-point in 1990s)',
     xlab = 'Season', ylab = 'Total Team Fouls', col = 'black', pch = 16)

points(top_10_foul_sum$Season[-48], top_10_foul_sum$Total_Fouls[-48], col = 'black', pch = 19)
abline(v = ci, col = 'blue', lwd = 2)
abline(v = mean_posterior, col = 'red', lwd = 2)

# Add legend
legend("topright", legend = c("Total Fouls", "95% Credible Interval", "Posterior Mean"),
       col = c("black", "blue", "red"), pch = c(19, NA, NA), lwd = 2, cex = 0.5,
       title = "Legend")

```

## Normal-Normal Model on Average Team Score:

1. First, load the team per game data and call the necessary functions to clean it:

```{r message=FALSE}

team_per_game_data = read_csv("data/team_per_game_stats.csv")

mean_points_per_game = get_team_average_points(team_per_game_data)

head(mean_points_per_game)

```

2. Plot the data:

```{r}

# Create the histogram with correct title
ggplot(data = team_per_game_data, aes(x = PTS)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Average Team Score",
    x = "Team Score Per Game",
    y = "Frequency"
  ) +
  theme_minimal()

```


```{r}

ggplot(mean_points_per_game, aes(x = as.factor(Season), y = AVG_PTS)) +
  geom_point(color = "blue", size = 2) + 
  labs(
    title = "Average Team Score By Season",
    x = "Season",
    y = "Average Team Score",
    caption = "Data Source: BBR"
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

3. Run the models:

1) Assuming there is only one change-point:

```{r}

result = run_normal_model_one_changepoint(mean_points_per_game$AVG_PTS, alpha = 0.01, beta = 0.01, sigma_theta = 10)

ci = HPDinterval(as.mcmc(result + 1977))

# Plot histogram of potential change-points:
hist(result + 1977, breaks = 50, col = "skyblue", border = "black",
     main = "Histogram of Potential Changepoints",
     xlab = "Potential Change-point Season",
     ylab = "Frequency")

# Add the credible intervals and the corresponding legend:
abline(v = ci, lwd = 3, col = 'blue')
legend("topright", legend = "95% Credible Interval", col = "blue", lwd = 3, cex = 0.5)

```

```{r}

ci = HPDinterval(as.mcmc(result + 1977))
mean_posterior = round(mean(result) + 1977)

plot(mean_points_per_game$Season[-48], mean_points_per_game$AVG_PTS[-48],
     main = 'Average Team Score by Season (Change-point in Early 1990s)',
     xlab = 'Season', ylab = 'Average Team Score', col = 'black', pch = 16)

points(mean_points_per_game$Season[-48], mean_points_per_game$AVG_PTS[-48], col = 'black', pch = 19)
abline(v = ci, col = 'blue', lwd = 2)
abline(v = mean_posterior, col = 'red', lwd = 2)

# Add legend
legend("topright", legend = c("Average Team Score", "95% Credible Interval", "Posterior Mean"),
       col = c("black", "blue", "red"), pch = c(19, NA, NA), lwd = 2, cex = 0.5,
       title = "Legend")

```


2) Assuming there are two change-points:

```{r}

result = run_normal_model_two_changepoints(mean_points_per_game$AVG_PTS, alpha = 0.01, beta = 0.01, sigma_theta = 10)

```

```{r}

posterior_changepoint1 = result[[1]]
posterior_changepoint2 = result[[2]]

ci = HPDinterval(as.mcmc(posterior_changepoint1 + 1977))

# Plot histogram of potential change-points:
hist(posterior_changepoint1 + 1977, breaks = 50, col = "skyblue", border = "black",
     main = "Histogram of Potential First Change-point",
     xlab = "Potential Change-point Season",
     ylab = "Frequency")

# Add the credible intervals and the corresponding legend:
abline(v = ci, lwd = 3, col = 'blue')
legend("topright", legend = "95% Credible Interval", col = "blue", lwd = 3, cex = 0.5)

```

```{r}

ci = HPDinterval(as.mcmc(posterior_changepoint2 + 1977))

# Plot histogram of potential change-points:
hist(posterior_changepoint2 + 1977, breaks = 50, col = "skyblue", border = "black",
     main = "Histogram of Potential Second Change-point",
     xlab = "Potential Change-point Season",
     ylab = "Frequency")

# Add the credible intervals and the corresponding legend:
abline(v = ci, lwd = 3, col = 'blue')
legend("topright", legend = "95% Credible Interval", col = "blue", lwd = 3, cex = 0.5)

```


```{r}

ci_1 = HPDinterval(as.mcmc(posterior_changepoint1 + 1977))
ci_2 = HPDinterval(as.mcmc(posterior_changepoint2 + 1977))

# Create the plot
plot(mean_points_per_game$Season, mean_points_per_game$AVG_PTS,
     main = 'Average Team Score by Season (Two Change-points)',
     xlab = 'Season', ylab = 'Average Team Score',
     col = 'black', pch = 16)

# Add vertical lines for credible intervals and posterior means for change point 1
abline(v = ci_1, lwd = 2, col = 'blue')
abline(v = mean(posterior_changepoint1) + 1977, lwd = 2, lty = 2, col = 'purple')

# Add vertical lines for credible intervals and posterior means for change point 2
abline(v = ci_2, lwd = 2, col = 'red')
abline(v = mean(posterior_changepoint2) + 1977, lwd = 2, lty = 2, col = 'yellow')

# Add smaller legend
legend("topright", legend = c("Change Point 1 CI", "Change Point 1 Mean", "Change Point 2 CI", "Change Point 2 Mean"),
       col = c("blue", "purple", "red", "yellow"), lwd = 2, lty = c(1, 2, 1, 2), cex = 0.4,
       title = "Legend")

```


4. Model diagnostics:

```{r}

plot(as.ts(posterior_changepoint1))

```

```{r}

plot(as.ts(posterior_changepoint2))

```

```{r}

effectiveSize(posterior_changepoint1); effectiveSize(posterior_changepoint2)

```

## Normal-Normal Model on Team Pace

```{r warning=FALSE}

team_advanced_data = read_csv("data/team_advanced_stats.csv")

average_team_pace = get_team_average_pace(team_advanced_data)

head(average_team_pace)

```

```{r}

ggplot(average_team_pace, aes(x = as.factor(Season), y = AVG_Pace)) +
  geom_point(color = "blue", size = 2) + 
  labs(
    title = "Average Team Pace By Season",
    x = "Season",
    y = "Average Team Pace",
    caption = "Data Source: BBR"
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


One change-point Model:

```{r}

result = run_normal_model_one_changepoint(average_team_pace$AVG_Pace, alpha = 0.01, beta = 0.01, sigma_theta = 10)

ci = HPDinterval(as.mcmc(result + 1977))

# Plot histogram of potential change-points:
hist(result + 1977, breaks = 50, col = "skyblue", border = "black",
     main = "Histogram of Potential Changepoints",
     xlab = "Potential Change-point Season",
     ylab = "Frequency")

# Add the credible intervals and the corresponding legend:
abline(v = ci, lwd = 3, col = 'blue')
legend("topright", legend = "95% Credible Interval", col = "blue", lwd = 3, cex = 0.5)

```

```{r}

ci = HPDinterval(as.mcmc(result + 1977))
mean_posterior = round(mean(result) + 1977)

plot(average_team_pace$Season[-48], average_team_pace$AVG_Pace[-48],
     main = 'Average Team Pace by Season (Change-point in Early 1990s)',
     xlab = 'Season', ylab = 'Average Team Pace', col = 'black', pch = 16)

points(average_team_pace$Season[-48], average_team_pace$AVG_Pace[-48], col = 'black', pch = 19)
abline(v = ci, col = 'blue', lwd = 2)
abline(v = mean_posterior, col = 'red', lwd = 2)

# Add legend
legend("topright", legend = c("Average Team Pace", "95% Credible Interval", "Posterior Mean"),
       col = c("black", "blue", "red"), pch = c(19, NA, NA), lwd = 2, cex = 0.5,
       title = "Legend")

```


Two change-points model:

```{r}

result = run_normal_model_two_changepoints(average_team_pace$AVG_Pace, alpha = 0.01, beta = 0.01, sigma_theta = 10)

```

Histograms of change-points:

```{r}

posterior_changepoint1 = result[[1]]
posterior_changepoint2 = result[[2]]

ci = HPDinterval(as.mcmc(posterior_changepoint1 + 1977))

# Plot histogram of potential change-points:
hist(posterior_changepoint1 + 1977, breaks = 50, col = "skyblue", border = "black",
     main = "Histogram of Potential First Change-point",
     xlab = "Potential Change-point Season",
     ylab = "Frequency")

# Add the credible intervals and the corresponding legend:
abline(v = ci, lwd = 3, col = 'blue')
legend("topright", legend = "95% Credible Interval", col = "blue", lwd = 3, cex = 0.5)

```


```{r}

ci = HPDinterval(as.mcmc(posterior_changepoint2 + 1977))

# Plot histogram of potential change-points:
hist(posterior_changepoint2 + 1977, breaks = 50, col = "skyblue", border = "black",
     main = "Histogram of Potential Second Change-point",
     xlab = "Potential Change-point Season",
     ylab = "Frequency")

# Add the credible intervals and the corresponding legend:
abline(v = ci, lwd = 3, col = 'blue')
legend("topright", legend = "95% Credible Interval", col = "blue", lwd = 3, cex = 0.5)

```


```{r}

posterior_changepoint1 = result[[1]]
posterior_changepoint2 = result[[2]]

ci_1 = HPDinterval(as.mcmc(posterior_changepoint1 + 1977))
ci_2 = HPDinterval(as.mcmc(posterior_changepoint2 + 1977))

# Create the plot
plot(average_team_pace$Season, average_team_pace$AVG_Pace,
     main = 'Average Team Pace by Season (Two Change-points)',
     xlab = 'Season', ylab = 'Average Team Pace',
     col = 'black', pch = 16)

# Add vertical lines for credible intervals and posterior means for change point 1
abline(v = ci_1, lwd = 2, col = 'blue')
abline(v = mean(posterior_changepoint1) + 1977, lwd = 2, lty = 2, col = 'purple')

# Add vertical lines for credible intervals and posterior means for change point 2
abline(v = ci_2, lwd = 2, col = 'red')
abline(v = mean(posterior_changepoint2) + 1977, lwd = 2, lty = 2, col = 'yellow')

# Add smaller legend
legend("topright", legend = c("Change Point 1 CI", "Change Point 1 Mean", "Change Point 2 CI", "Change Point 2 Mean"),
       col = c("blue", "purple", "red", "yellow"), lwd = 2, lty = c(1, 2, 1, 2), cex = 0.4,
       title = "Legend")

```


Next steps:
1. Offensive/defensive ratings
2. FT% (prove the results from previous works)
